---
title: 'Analyzing Potential Correlation Between Employment and Violent Crime Rate (2010 to 2020)'
author: "Ian Lacy"
date: "2023-12-01"
output:
  html_document: default
  fig_width: "auto"
  fig_height: "auto"
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

suppressMessages(library(readr))
suppressMessages(library(ggplot2))
suppressMessages(library(caret))
suppressMessages(library(tidyr))
suppressMessages(library(dplyr))
suppressMessages(library(reshape2))



c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

cbbPalette1 <-  c("#0072B2","#D55E00" )
cbbPalette2 <-  c("#D55E00", "#0072B2")
```



## Introduction

Some research has indicated that lower employment rates may be linked to higher rates of violent crime (UC Davis,2022).
According to Anirban Basu,chairman and chief executive of Sage Policy Group, an economic and policy consulting firm headquartered in Baltimore, “The private sector in Maryland is largely stagnant.” Basu posited that Maryland’s low unemployment rate is likely not because of a competitive labor market, but that people have moved elsewhere for other opportunities. As Maryland’s economic numbers have stagnated, there has also been a rise in violent crime. “Homicides in the state rose to a ten-year high of 573 in 2020, a nearly 35% increase from 2010. The largest year-to-year increase occurred in 2015 when homicides rose to 553, a 52% increase from the previous year,” according to reporting from The Maryland Daily Record. 

Given that the private sector is lagging and violent crime is surging, I decided to look at county-by-county data to see if employment rates are correlated with increases in violent crime. I will do this by visualizing publicly available data and seeing if there are any noticeable correlations. 

## Materials and Methods

To conduct this research I sourced county employment and crime data using data from two datasets from the Maryland Open Data Portal (opendata.maryland.gov). These datasets included:

* Violent Crime & Property Crime by County: 1975 to Present
  + This data is derived from the Maryland State Police's annual Uniform Crime Reports.
  + This data provides information by year, county, and crime type. These crime types were further divided by property crime (mostly types of theft) and violent crime (assaults). This is a simplified overview of the data, so I encourage the reader of this document to look at the original data for further information.
* Maryland and its Jurisdictions Full and Part Time Jobs by NAICS: 2010-2020 Historical and 2025-2050 Projected 
  + This data is provided by Maryland's Department of Planning.
  + This data provides information by year, county, and job sector.
  
  
I then imported this data into RStudio using R version 4.3.0 (2023-04-21 ucrt) -- "Already Tomorrow."

Throughout my analysis I used a variety of libraries to transform and visualize the data. These libraries and the steps I took are listed throughout the document.

## Results

### Exploratory Analysis

As part of the analysis process I conducted some basic data cleaning and produced some basic visualizations to help figure out how to proceed with my analysis. 

Data cleaning revealed that there were a lot of missing data in the employment information by job sector. This is probably due to certain industries being over/under represented or not present in certain counties based on real world factors like the presence of farms or proximity to resource extraction industries. Overall, there were more than 800 pieces of NA data,  so I decided to just use total employment numbers by county from 2010 to 2020. I  then decided to see if there were any outliers in the data to see if any counties need to be dropped from the analysis because they may skew the data. 

```{r}
suppressMessages(MD_Jobs <- read_csv("C:/Users/ianla.000/Downloads/MD_jobs.csv"))
suppressMessages(PopData <- read_csv("C:/Users/ianla.000/Downloads/Pop Data.csv"))

```

```{r}
summary(MD_Jobs)
```
```{r}
suppressMessages(MD_Employ <- read_csv("C:/Users/ianla.000/Downloads/Total_Employ_MD.csv"))

MD_Employ$EmploymentRatePer100K <- (MD_Employ$Employed / MD_Employ$Population) * 100000

MD_Employ$EmploymentRatePer10k <- (MD_Employ$Employed / MD_Employ$Population) * 100

MD_Employ$EmploymentRatePer10k <- (MD_Employ$EmploymentRatePer10k/100) *10000

MD_Employ

```


```{r}
ggplot(MD_Employ, aes(x = County, y = Employed, fill = County)) +
  geom_boxplot() +
  labs(title = "Distribution of Employed People by County from 2010 to 2020",
       x = "County",
       y = "Number of Employed People") +
   scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

Looking at the above graph we see that Anne Arundel County,Baltimore County, Howard County, Montgomery, and Prince George's county are appear to be significant outliers in the number of total people employed. This makes sense because these are the largest counties by population in the state. This may make it difficult to determine how total employment affects crime rates unless we adjust for population, which the crime data does do by adjusting for crime per 100,000 people. Therefore, I created a feature that  adjusts employment to a per capita measurement per 100k residents.


Moving on, I did a similar plot to look at the total crime (not adjusted by population) from year to year by county. 

```{r}
suppressMessages(MD_Crime <- read_csv("C:/Users/ianla.000/Downloads/Crime_total.csv"))
```

```{r}
#summary(MD_Crime)
MD_Crime
```

```{r}
ggplot(MD_Crime, aes(x = County, y = Total, fill = County)) +
  geom_boxplot() +
  labs(title = "Distribution of Total Crimes by County from 2010 to 2020",
       x = "County",
       y = "Total Number of Crimes") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
In general we see a similar trend to the employment data, with higher total crime numbers relative to the population. This confirms that using adjusted metrics for crime rates per 100,00 will be necessary to avoid skewed data.

Next, because there are so many types of violent crime and property crime, I thought it would be better to look at the total violent crime and property crime rates over time to see if there are any overall trends across the different counties.

```{r}
suppressMessages(MD_Crime100 <- read_csv("C:/Users/ianla.000/Downloads/PerCapitaCrime_MD.csv"))
```

```{r}
# Reshape data to long format
library(tidyr)
MD_Crime_100_long <- pivot_longer(MD_Crime100, cols = c(ViolentCrime, PropertyCrime), names_to = "CrimeType", values_to = "CrimeRate")
MD_Crime_100_long
```



```{r}
  ggplot(filter(MD_Crime_100_long, County %in% c("Allegany County","Anne Arundel County","Baltimore City", "Baltimore County", "Calvert County","Caroline County")), 
    aes(x = Year, y = CrimeRate, fill = CrimeType)) +
    geom_bar(stat="identity") +
    facet_wrap(County ~ ., scales = "free_y") +
    labs(title = "Property and Violent Crime by County (2010-2020)",
       x = "Year",
       y = "Number of Crimes Committed") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette1, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
  ggplot(filter(MD_Crime_100_long, County %in% c("Carroll County","Cecil County","Charles County", "Dorchester County", "Frederick County","Garrett County")), 
    aes(x = Year, y = CrimeRate, fill = CrimeType)) +
    geom_bar(stat="identity") +
    facet_wrap(County ~ ., scales = "free_y") +
    labs(title = "Property and Violent Crime by County (2010-2020)",
       x = "Year",
       y = "Number of Crimes Committed") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette1, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
  ggplot(filter(MD_Crime_100_long, County %in% c("Harford County","Howard County","Kent County", "Montgomery County", "Prince George's County","Queen Anne's County")), 
    aes(x = Year, y = CrimeRate, fill = CrimeType)) +
    geom_bar(stat="identity") +
    facet_wrap(County ~ ., scales = "free_y") +
    labs(title = "Property and Violent Crime by County (2010-2020)",
       x = "Year",
       y = "Number of Crimes Committed") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette1, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
  ggplot(filter(MD_Crime_100_long, County %in% c("Somerset County","St. Mary's County","Talbot County", "Washington County", "Wicomico County","Worcester County")), 
    aes(x = Year, y = CrimeRate, fill = CrimeType)) +
    geom_bar(stat="identity") +
    facet_wrap(County ~ ., scales = "free_y") +
    labs(title = "Property and Violent Crime by County (2010-2020)",
       x = "Year",
       y = "Number of Crimes Committed") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette1, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
After looking at the breakdown of property crime versus violent crime I realized several things: 1) property crime is by far the most prevalent form of crime in any Maryland county. 2) Overall, violent crime rates seem relatively flat over time. 3) Property crime rates appear to consistently be falling over time.

Based on this, I decided to look at the distributions of property and violent crime rates adjusted for population. All rates are presented as per 100k residents.

```{r}
ggplot(MD_Crime100, aes(x = County, y = PropertyCrime, fill = County)) +
  geom_boxplot() +
  labs(title = "Property Crimes per 100k Residents by County from 2010 to 2020",
       x = "County",
       y = "Property Crime Rates") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
```{r}
ggplot(MD_Crime100, aes(x = County, y = ViolentCrime, fill = County)) +
  geom_boxplot() +
  labs(title = "Property Crimes per 100k Residents by County from 2010 to 2020",
       x = "County",
       y = "Property Crime Rates") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
















































