---
title: "Homework: Analyzing Count Data"
author: "Data Science for Bioinformatics S2024"
output: word_document
---

Student Name: Ian Lacy


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


## Reviewing `SummarizedExperiment`

The `SummarizedExperiment` data structure is commonly used to provide data for a related series of high-throughput experiments.

One way to familiarize yourself with this data structure is to create a new one:

```{r}
set.seed(123)
# Create a toy SummarizedExperiment object
counts <- matrix(rnbinom(150, mu=50, size=10), nrow=15, ncol=10)
rownames(counts) <- paste0("gene", 1:15)
colnames(counts) <- paste0("sample", 1:10)
col_data <- data.frame(condition=rep(c("A", "B"), each=5))
row_data <- data.frame(symbol=paste0("gene", 1:15))
```

Let's take a look at these datastructures:

```{r}
knitr::kable(counts)
```

```{r}
col_data
```

```{r}
row_data
```


We are now ready to create a `SummarizedExperiment` object like so:

```{r}
library(SummarizedExperiment)
se <- SummarizedExperiment(assays=list(counts=counts), colData=col_data, rowData=row_data)
```

## Retrieving Data

We can retrieve the count data from `SummarizeExperiments` with function assays:

```{r}
knitr::kable(assays(se)[[1]])
```

Similarly, we can retriee the column data and row-data with `colData` and `rowData` respectively:

```{r}
colData(se)
```


```{r
rpwData(se)}
```

## Using the 'Airway' RNA-Seq Data

The airway package contains a dataset of gene expression values from a study of airway epithelial cells in response to glucocorticoid treatment. This dataset is based on RNA-Seq data and includes information on gene expression levels for thousands of genes across several experimental conditions.


```{r}
BiocManager::install('airway')
library(airway)
data(airway)
class(airway)
```



## Inspecting Read Match Count Data

### Column Meta-Data

```{r}
colData(airway)
```


### Count Matrix

```{r}
counts <- assays(airway)$counts
counts <- counts[1:30,] # no more than this many genes
knitr::kable(counts)
```

## Our DIY Approach

Before utilizing ready-made approaches for analyzing RNA-Seq data, we will use our own approach(es). This will help us understand what actually the challenge is when analyzing high-throughput experiment data.

Our approach will be to view RNA-Seq data as *count* data where we count for each gene or for each transcript how many reads we find in our sample that align with a reference sequence from a database corresponding to this transcript.

Counts of random events are often well-described by the Poisson distribution, so this is how we will start.

## 'Melting' The Matrix

Before fitting models to the count data, we have to convert it to a 'melted form' like so:

```{r}
library(reshape2) # for 'melt'
counts_df <- as.data.frame(counts)
countsm = melt(t(counts_df))  # little trick to swap columns and rows
colnames(countsm) <- c('Sample', 'Gene', 'Count')
knitr::kable(head(countsm,n=10))
```

## Modeling RNA-Seq Count Data Assuming a Poisson Distribution

We can create a model of the count data using a generalized linear model (glm). This means that 'under the hood' R uses still code for the regular linear model (function `lm`), but the input data is transformed before starting a regular lineary regresion so that we can treat it as if we deal with a new way of fitting specialized data that is not linear. The code is showing below.

```{r}
# Fit a Poisson GLM to the count data
poisson_model <- glm(Count ~ Sample+Gene, data = countsm, family = "poisson")
summary(poisson_model)
```

Let's take a look at the obtained summary of the model. 
Most prominent is the section called 'Coefficients'. We will focus on their name (left-most column), the column 'Estimate' and the P-value ("Pr(>|z|)"). 

For a regular linear regression, the column 'Estimate' corresponds to the slope of a line. Because we performed a specialized type of regression to fit counts of a Poisson distribution, the column 'Estimate' has to be interpret differently. It stands for the logarithm of the predicted outcome for a that variable to be increased by one unit. In our case this stands for the predicted count values of a certain gene divided by the predicted count values averages over all except that particular gene. For example, for gene ENSG00000000971, the summary shows a value for the Estimate of 2.096109. This means $\log(\frac{count_g}{count_{ng}})=2.096$ or $count_g=count_{ng}e^{2.096}\approx 8.13 * count_{ng}$ 
In other words, the gene ENSG00000000971 has count values that are about 8 times higher compared to the count values of all the other genes.
We should also look at the P-values and ensure that they indicate statistical significance. 


## Exercise 1

a) Interpret the summary of the model fit. What is the maximum and minimum deviation between the ground-truth and prediction? Hint: Look at min, max of Deviance Residuals

```{r}
res <- residuals(poisson_model, type = "deviance")
summary(res)

```
The maximum deviation is 29.35 and the minimum deviation is -27.04.


b) Consider gene ENSG00000000938. What is the raw Estimate value shown in the summary for this gene? Does this indicate that the count values for this genes are higher or lower compared to the counts of the other genes? 
The raw estimate for gene ENSG00000000938 is -7.590010. This means that the count values for this gene are much lower compared to the counts of other genes.  

c) Based on what you obtained in b), what is the odds ratio in terms of gene counts for this gene divided by the average counts of all the other genes?
Given that estimates determine the relationship between the input and the logarithmic output, using the formula e^-7590010 the result would be approximately .00051. 

d) Apply the function `tidy` from R package `broom` to the Poisson model. Use function `knitr::kable` to print the resulting data frame. 

```{r}
if (!require(broom)) {
  install.packages("broom")
}
```

```{r}
knitr::kable(broom::tidy(poisson_model))
```



## Modeling RNA-Seq Count Data Assuming a Negative Bionomial Distribution

Next, we will use a similar approach for fitting a model that assumes a negative binomial distribution of the count data. One can view the Poisson distribution as a special case of the negative binomial distribution where the variance is given through the mean values. In contrast, the negative binomial distribution does not have this restriction, it can allow for larger variances as they are common in biological data.

The function for fitting a generalized linear model based on this distribution can be found in the MASS package (the name of the library is based on the book title of the classic "Modern Applied Statistics with S").

```{r}
library(MASS) # for glm.nb
# Fit a negative binomial GLM to the count data
nb_model <- glm.nb(Count ~ Gene + Sample, data = countsm)
summary(nb_model)
```


## Exercise 2: Comparing Residuals of the Models

Let us inspect the quality of the generated models. One way is to look at the difference between the ground-truth values and the predicted values. This is called 'residual' in statisics.

The formulate is: residual = observed y - predicted y

a) Use function `residuals` to compute how much our predictions deviate from the ground-truth values of the counts for the Poisson model. Summarize the numeric results (be creative).

```{r}
poi_res <- residuals(poisson_model)
summary(poi_res)
```

b) Apply the same approach to the residuals corresponding to the negative binomial model.
```{r}
nb_res <- residuals(nb_model)
summary(nb_res)

```

c) Find a way to plot your findings from a) and b) in an informative way. Be creative: you can create a combined plot or two separate plots; feel free to use plotting functions from either base-R or other libraries such as ggplot.
```{r}
residual_combo <- data.frame("Negative Binomal" = poi_residuals, "Poisson" = nb_res)
residual_combo
combined_melt <- melt(residual_combo)
combined_melt
```

```{r}
library(ggplot2)
```


```{r}
ggplot(combined_melt,aes(x = value,fill = variable)) +
geom_histogram(alpha=0.5, position = "identity") +
ggtitle("Density Plot of Residuals by Model Type") +
xlab("Residuals") + ylab("Density")
```


d) Based on your findings in a)-c), what model seems superior and why?
The Poisson model is better. Based on the histogram we see that the residuals for the Poisson model are clustered around zero, indicating that the predictions made are close to the observed values.
The model also looks generally bell shaped, indicating normal distribution.


## Choosing the Better Model using the Aikike Information Criterion

In statistic and machine learning we have the challenge that models with more parameters are intrinsically better able to obtain better fits to data compared
to models with less parameters. But 'more' is not necessarily 'better' as it can lead to over-fitting! Somehow we need a way to balance the need for fitting the data well with the need to avoid a proliferation in model complexity.

The Akaike Information Criterion (AIC) is a measure of the quality of a statistical model that takes into account both how well the model fits the data and how complex the model is. The AIC score is calculated by taking the log-likelihood of the model and subtracting a penalty term that increases as the number of parameters in the model increases.

The AIC helps us choose the best model by balancing two things: how well the model fits the data and how complicated the model is. The AIC score is lower for models that fit the data well while using fewer parameters, which means they are more likely to generalize well to new data. So, the model with the lowest AIC score is usually considered the best model for a given set of data.

In R we have the function `AIC` that computes this metric for a given model (for example `ACI(model)` )

## Exercise 3

a) Use the function `AIC()` to compute the AIC metric for the model corresponding to the Poisson distribution. What is the value?

```{r}
AIC_pois <- AIC(poisson_model)
AIC_pois
```

b) Use the function `AIC()` to compute the AIC metric for the model corresponding to the negative binomial distribution. What is the value?

```{r}
AIC_binomial <- AIC(nb_model)
AIC_binomial
```

c) Based on your results in a) and b), which model seems to be preferrable?
Based on the above scores, the nb_model appears to be better because the score is lower. 


## Exercise 4: General understanding of Count Matrix Analysis

a) Use common sense: Why is the number RNA-Seq reads one identifies for a certain transcript variant approximately proportional to the RNA sequence length of that transcript variant?
The longer a sequence is, the more likely it is that there will be more reads aligned to it.

b) It is tempting to use the odds-ratios of counts we obtained earlier in form of the `glm` estimates as measures for gene expression. Why is that not directly workable? Hint: see a)

A longer transcript is more likely to have more reads which could increase the appearance of expression but this is only due to the size of the transcript. A shorter transcript might show lower expression. To be workable, you would need to normalize the ratios for them to be useful.


c) Another key aspect that our initial modeling using `glm` did not account for is that experiments are typically grouped together, and one compares one condition versus another, each supported by counts for several samples. Present ideas how our models could be refined such that we do not simply compare each sample individually but instead of sample groups by different treatments.

You could group things using group-level predictor which would create a categorical variable that would estimate based on that variable. An example could be adding "tissue type" which would delineate samples by tissue type. We could also include interaction terms to see how that influences counts.


## Resources

* <https://stats.oarc.ucla.edu/r/dae/poisson-regression/>
* <https://bioconductor.org/packages/release/data/experiment/html/airway.html>

