---
title: 'Term Project'
author: "Ian Lacy"
date: "2023-12-01"
output:
  html_document: default
  fig_width: "auto"
  fig_height: "auto"
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(reticulate)
```

```{r}
suppressMessages(library(readr))
suppressMessages(library(ggplot2))
suppressMessages(library(caret))
suppressMessages(library(tidyr))
suppressMessages(library(dplyr))
suppressMessages(library(reshape2))



c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown"
)

cbbPalette1 <-  c("#D55E00", "#0072B2")
```
```
# Cleaning Data


```{r}

```


## Cleaning Data

In this section I did an exploratory analysis of the datasets to see if they would be useful for the project. I did this by cleaning the data like converting missing data to NA's and looking at the distribution of data to see if there are any outliers.

```{r}
suppressMessages(MD_Jobs <- read_csv("C:/Users/ianla.000/Downloads/MD_jobs.csv"))
suppressMessages(PopData <- read_csv("C:/Users/ianla.000/Downloads/Pop Data.csv"))

```

```{r}
summary(MD_Jobs)
```
One thing that immediately stood out to me was the large number of NA's in each column of data. This equated to over 800 pieces of missing data, leaving large gaps in the data. This makes evaluating by sector quite difficult, so I decided to just use total employment numbers by county from 2010 to 2020. I decided to see if there are any outliers in the data to see if any counties need to be dropped from the analysis because they may skew the data.  


```{r}
suppressMessages(MD_Employ <- read_csv("C:/Users/ianla.000/Downloads/Total_Employ_MD.csv"))
MD_Employ
```


## Exploratory Anaylsis
```{r}
#MD_Employ
MD_Employ$EmploymentRatePer100K <- (MD_Employ$Employed / MD_Employ$Population) * 100000

MD_Employ$EmploymentRatePer10k <- (MD_Employ$Employed / MD_Employ$Population) * 100

MD_Employ$EmploymentRatePer10k <- (MD_Employ$EmploymentRatePer10k/100) *10000

MD_Employ

```

I added a feature that adjusts employment to a per capita measurement per 100k residents. 

```{r}
ggplot(MD_Employ, aes(x = County, y = Employed, fill = County)) +
  geom_boxplot() +
  labs(title = "Distribution of Employed People by County from 2010 to 2020",
       x = "County",
       y = "Number of Employed People") +
   scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```
Looking at the above graph we see that Anne Arundel County,Baltimore County, Howard County, Montgomery, and Prince George's county are appear to be significant outliers in the number of total people employed. This makes sense because these are the largest counties by population in the state. This may make it difficult to determine how total employment affects crime rates unless we adjust for population, which the crime data does do by adjusting for crime per 100,000 people.


Moving on, I did a similar plot to look at the total crime from year to year by county. 

```{r}



ggplot(MD_Employ, aes(x = Year, y = EmploymentRatePer100K, color = County)) +
  geom_line() +
  geom_point()+
  labs(title = "Employment per 100k by County from 2010 to 2020",
       x = "Year",
       y = "Number of People Employed per 100k") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = c25)

```
Looking at the above graph we see that +6Anne Arundel County,Baltimore County, Howard County, Montgomery, and Prince George's county are appear to be significant outliers in the number of total people employed. This makes sense because these are the largest counties by population in the state. This may make it difficult to determine how total employment affects crime rates unless we adjust for population, which the crime data does do by adjusting for crime per 100,000 people.


```{r}



ggplot(PopData, aes(x = Year, y = ViolentRate, color = County))+
  geom_line() +
  geom_point()+
  labs(title = "Violent Crime Rate per 100k by County from 2010 to 2020",
       x = "Year",
       y = "Number of Crimes per 100k") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = c25)

```



Moving on, I did a similar plot to look at the total crime from year to year by county. 

```{r}
suppressMessages(MD_Crime <- read_csv("C:/Users/ianla.000/Downloads/Crime_total.csv"))
```

```{r}
#summary(MD_Crime)
#MD_Crime
```

```{r}
ggplot(MD_Crime, aes(x = County, y = Total, fill = County)) +
  geom_boxplot() +
  labs(title = "Distribution of Total Crimes by County from 2010 to 2020",
       x = "County",
       y = "Total Number of Crimes") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
In general we see a similar trend to the employment data, with higher total crime numbers relative to the population. This confirms that using adjusted metrics for crime rates per 100,00 will be necessary to avoid skewed data.

Next, because there are so many types of violent crime and property crime, I thought it would be better to look at the total violent crime and property crime rates over time to see if there are any overall trends across the different counties.

```{r}
suppressMessages(MD_Crime_100 <- read_csv("C:/Users/ianla.000/Downloads/Crime_A-B.csv"))
MD_Crime_100 

```

```{r}
# Reshape data to long format
library(tidyr)
MD_Crime_100_long <- pivot_longer(MD_Crime_100, cols = c(ViolentCrime, PropertyCrime), names_to = "CrimeType", values_to = "CrimeRate")
#MD_Crime_100_long
```

```{r}
cbbPalette <-  c("#0072B2", "#D55E00")
# Create a plot with different panels for each county
ggplot(MD_Crime_100_long, aes(x = Year, y = CrimeRate, fill = CrimeType)) +
  geom_bar(stat="identity") +
  facet_wrap(County ~ ., scales = "free_y") +
  labs(title = "Property and Violent Crime by County (2010-2020)",
       x = "Year",
       y = "Number of Crimes Committed") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
suppressMessages(MD_Crime_2 <- read_csv("C:/Users/ianla.000/Downloads/Crime_2.csv"))
MD_Crime_2 

```

```{r}
# Reshape data to long format
library(tidyr)
MD_Crime_2_long <- pivot_longer(MD_Crime_2, cols = c(ViolentCrime, PropertyCrime), names_to = "CrimeType", values_to = "CrimeRate")
MD_Crime_2_long
```

```{r}
# Create a plot with different panels for each county
ggplot(MD_Crime_2_long, aes(x = Year, y = CrimeRate, fill = CrimeType)) +
  geom_bar(stat="identity") +
  facet_wrap(County ~ ., scales = "free_y") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
suppressMessages(MD_Crime_3 <- read_csv("C:/Users/ianla.000/Downloads/Crime3.csv"))
#MD_Crime_3 

```

```{r}
# Reshape data to long format
library(tidyr)
MD_Crime_3_long <- pivot_longer(MD_Crime_3, cols = c(ViolentCrime, PropertyCrime), names_to = "CrimeType", values_to = "CrimeRate")
#MD_Crime_3_long
```

```{r}
# Create a plot with different panels for each county
ggplot(MD_Crime_3_long, aes(x = Year, y = CrimeRate, fill = CrimeType)) +
  geom_bar(stat="identity") +
  facet_wrap(County ~ ., scales = "free_y") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
suppressMessages(MD_Crime_4 <- read_csv("C:/Users/ianla.000/Downloads/Crime4.csv"))
#MD_Crime_3 

```

```{r}
# Reshape data to long format
library(tidyr)
MD_Crime_4_long <- pivot_longer(MD_Crime_4, cols = c(ViolentCrime, PropertyCrime), names_to = "CrimeType", values_to = "CrimeRate")
#MD_Crime_3_long
```

```{r}
# Create a plot with different panels for each county
ggplot(MD_Crime_4_long, aes(x = Year, y = CrimeRate, fill = CrimeType)) +
  geom_bar(stat="identity") +
  facet_wrap(County ~ ., scales = "free_y") +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_fill_manual(values = cbbPalette, name = "Crime Type", labels = c("Property Crime", "Violent Crime")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

After looking at the breakdown of property crime versus violent crime I realized several things: 1) property crime is by far the most prevalent form of crime in any Maryland county. 2) Overall, violent crime rates seem relatively flat over time. 3) Property crime rates appear to consistently be falling over time.

Based on this, I decided to look at the distributions of property and violent crime rates adjusted for population. All rates are presented as per 100k residents.

```{r}
suppressMessages(MD_Crime_larceny <- read_csv("C:/Users/ianla.000/Downloads/larceny.csv"))
#MD_Crime_larceny 


ggplot(MD_Crime_larceny, aes(x = County, y = Prop100, fill = County)) +
  geom_boxplot() +
  labs(title = "Property Crimes per 100k Residents by County from 2010 to 2020",
       x = "County",
       y = "Property Crime Rates") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```
Suprisingly,

```{r}
suppressMessages(MD_Crime_Assault <- read_csv("C:/Users/ianla.000/Downloads/AssaultPerCapita.csv"))
#MD_Crime_Assault 

```

```{r}
ggplot(MD_Crime_Assault, aes(x = County, y = Violent100, fill = County)) +
  geom_boxplot() +
  labs(title = "Violent Crimes per 100k Residents by County from 2010 to 2020",
       x = "County",
       y = "Violent Crime per 100k") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
ggplot(MD_Employ, aes(x = County, y = EmploymentRatePer100K, fill = County)) +
  geom_boxplot() +
  labs(title = "Number of Employed People per 100k Residnets by County from 2010 to 2020",
       x = "County",
       y = "Number of Employed People per 100k") +
  scale_fill_manual(values = c25)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

```{r}
suppressMessages(PopData <- read_csv("C:/Users/ianla.000/Downloads/Pop Data.csv"))
```

```{r}
PopData
```
```{r}

PopData_long <- pivot_longer(PopData, cols = c(CrimeRate, EmploymentRate), names_to = "Type", values_to = "Rates")
PopData_long

```


```{r}
ggplot(filter(PopData_long, County %in% c("Allegany County","Anne Arundel County","Baltimore City", "Baltimore County", "Calvert County","Caroline County")), aes(x = Year, y = Rates)) +
  geom_line(aes(color = Type), size = .7) +
  facet_wrap(County ~ .)+
  labs(
    title = " Per Capita Employment and Crime Rates by County from 2010 to 2020",
    x = "Year",
    y = "Rates per 100k"
  ) +
  theme_minimal()+
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = cbbPalette1, name = "Crime Type", labels = c("Crime Rate", "Employment Rate")) +
  theme(
    legend.key.size = unit(0.3, "lines"),  # Adjust the size of the legend key
    legend.text = element_text(size = 8),   # Adjust the size of the legend text
  axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}


ggplot(filter(PopData_long, County %in% c("Carroll County","Cecil County","Charles County", "Dorchester County", "Frederick County","Garrett County")), aes(x = Year, y = Rates)) +
  geom_line(aes(color = Type), size = .7) +
  facet_wrap(County ~ .)+
  labs(
    title = " Per Capita Employment and Crime Rates by County from 2010 to 2020",
    x = "Year",
    y = "Rates per 100k"
  ) +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = cbbPalette1, name = "Crime Type", labels = c("Crime Rate", "Employment Rate")) + 
  theme(
    legend.key.size = unit(0.3, "lines"),  # Adjust the size of the legend key
    legend.text = element_text(size = 8),   # Adjust the size of the legend text
  axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

```

```{r}
ggplot(filter(PopData_long, County %in% c("Harford County","Howard County","Kent County", "Montgomery County", "Prince George's County","Queen Anne's County")), aes(x = Year, y = Rates)) +
  geom_line(aes(color = Type), size = .7) +
  facet_wrap(County ~ .)+
  labs(
    title = " Per Capita Employment and Crime Rates by County from 2010 to 2020",
    x = "Year",
    y = "Rates per 100k"
  ) +
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = cbbPalette1, name = "Crime Type", labels = c("Crime Rate", "Employment Rate")) +
    theme(
    legend.key.size = unit(0.3, "lines"),  # Adjust the size of the legend key
    legend.text = element_text(size = 8),   # Adjust the size of the legend text
  axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r}
ggplot(filter(PopData_long, County %in% c("Somerset County","St. Mary's County","Talbot County", "Washington County", "Wicomico County","Worcester County")), aes(x = Year, y = Rates)) +
  geom_line(aes(color = Type), size = .7) +
  facet_wrap(County ~ .)+
  labs(
    title = " Per Capita Employment and Crime Rates by County from 2010 to 2020",
    x = "Year",
    y = "Rates per 100k"
  ) +
  theme_minimal()+
  scale_x_continuous(limits = c(2010, 2020), breaks = seq(2010, 2020, 2))+
  scale_color_manual(values = cbbPalette1, name = "Crime Type", labels = c("Crime Rate", "Employment Rate")) +
  theme(
    legend.key.size = unit(0.3, "lines"),  # Adjust the size of the legend key
    legend.text = element_text(size = 8),   # Adjust the size of the legend text
  axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

