---
title: 'Assignment 2: Fix the Graphs'
author: "Ian Lacy"
date: "2023-10-24"
output:
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
matplotlib <- import("matplotlib")
matplotlib$use("Agg", force = TRUE)
```

```{r}
suppressMessages(library(readr))
suppressMessages(library(ggplot2))
suppressMessages(library(caret))
suppressMessages(library(tidyr))
suppressMessages(library(dplyr))
suppressMessages(library(reshape2))
suppressMessages(library(tidyr))
```

```{python}
import matplotlib as mpl
import matplotlib.pyplot as plt
import pandas as pd 

```



```{r}
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## Graph 1

```{r}
Percent <- c(62, 54, 54)
Party <- c("Democrats", "Republicans", "Independents")

PartyTotal <- data.frame(Party, Percent)
#PartyTotal

plot1 <- ggplot(PartyTotal, aes(x=Party, y=Percent, fill = Party)) + 
  geom_bar(stat="identity")+
 scale_fill_manual(name = "Party",values = cbbPalette, )+
  theme(plot.caption = element_text(size=8,hjust = 0))+
  ggtitle("Percent Who Agreed With Court by Party (2018)")+
  theme(plot.title = element_text(size=8,hjust=0.5))+
  theme(axis.title = element_text(size = 8))+
  theme(legend.title = element_text(size=8))

plot1 
```

> *The original graph was in black in white and the y-axis did not start at 0. Presenting the graph in black and white makes it difficult to immediately distinguish between the three data categories and not starting the graph at zero biases the data to make it look like Democrats overwhelmingly support the court compared to Republicans and Independents. This is not the case, as Democrats only support the court by eight percentage points. Another issue is that the graph shows percentages instead of the actual number of people.*
> *So what I did was make the graph color blind friendly and start the graph's y-axis at zero to show the relative difference between parties is not that great. I couldn't show absolute count data because it was not provided, so this the best I can do. I also added a caption to briefly explain the data.*

## Graph 2

```{r}
Deaths <- c(873,805,795,800,750,700,690,650,625,590,490,500,505,550,590,550,510,750,810,790,700,690,700,705,721)
Year <- c(1989,1990,1991,1992,1993,1994,1995,1996,1997,1998,1999,2000,2001,2002,2003,2004,2005,2006,2007,2008,2009,2010,2011,2012,2013
)

FLGunDeaths <- data.frame(Year, Deaths)
#FLGunDeaths

plot2 <- ggplot(FLGunDeaths, aes(x=Year, y=Deaths)) + 
  geom_area(fill = "#D55E00")+
  geom_point() +
  ggtitle("Florida Gun Deaths Per Year (1989 to 2013)") +
  xlab("Year") + 
  ylab("Deaths")+  
  theme(plot.title = element_text(size=14, hjust=0.5))+  
  theme(axis.title = element_text(size = 10))+
  theme(plot.caption = element_text(size=6,hjust = 0))

plot2
```

> *The original graph is inverted (the y-axis is flipped so that zero is at the top and higher numbers are near the bottom), making it appear that there is a dramatic drop in gun deaths after Florida passed the "Stand your Ground" law. In the caption it also says number of murders and not gun deaths. This misleading because murders are a specific crime whereas gun deaths are not.*
> *I fixed this by starting the y-axis at the bottom and used a colorblind friendly color to fill in the area under the line graph instead of the original red.*


## Graph 3

```{r}
Date <- c("March 27", "March 31")
Enrollees <- c(6000000, 7066000)

OBAMA <- data.frame(Date, Enrollees)
#OBAMA

plot3 <- ggplot(OBAMA, aes(x=Date, y=Enrollees)) + 
  geom_bar(stat = "identity", fill = "#0072B2" )+
  scale_y_continuous(labels = scales::label_number_si())+
  theme(plot.caption = element_text(size=8,hjust = 0))+
  ggtitle("Obamacare Enrollment (2014)")+
  theme(plot.title = element_text(size=12,hjust=0.5))+
  theme(axis.title = element_text(size = 10))+
  theme(legend.title = element_text(size=10))

plot3
```

> *The original graph was misleading because it did not accurately represent the two columns compared to one another. The March 27 column looked significantly smaller and not proportional to the March 31 graph. Also, the original did not have labels on  the y-axis, providing no frame of reference for viewers.*
> *I fixed this by introducing a y-axis that properly scaled the graph. I also changed the color of the bar charts to a color-blind-friendly choice.*


## Graph 4
```{python}

labels1='For', 'Against' 
sizes1 = [48, 52]

colors= CB_color_cycle = ['#377eb8', '#ff7f00', '#4daf4a',
                  '#f781bf', '#a65628', '#984ea3',
                  '#999999', '#e41a1c', '#dede00']                          
```

```{python}
explode = (0.05, 0.05)
B = fig, ax = plt.subplots()
ax.pie(sizes1, colors=colors, shadow = True, explode = explode, labels = None, autopct='%1.1f%%',pctdistance=1.3, startangle=90,radius=.8)
plt.legend(labels=labels1,loc= "lower left", fontsize="8")
plt.title('UKers For or Against Scottish Independence?', size=10, weight="bold") 
plt.show(B)
```

> *The original graph is just wrong. It's a pie chart that adds up to more than 100%. There was clearly an error in creating it. I'm assuming the creators meant to put 48% instead of 58%. In fact, it turns out that CNN actually completely misrepresented the data. The poll this graphic is based on actually found 52% against independence and 48 in favor: https://www.businessinsider.com/cnn-made-a-huge-mistake-on-a-scotland-independence-poll-2014-9.*
> *In line with the Business Insider article above, I presented a graphic that would appropriately visualize the data.*



## Graph 5
```{python}

labels='Yes', 'No', 'Neither' 
sizes = [50, 49, 1]
                          
CB_color_cycle = ['#377eb8', '#ff7f00', '#4daf4a',
                  '#f781bf', '#a65628', '#984ea3',
                  '#999999', '#e41a1c', '#dede00']    
                  
colors = CB_color_cycle

```


```{python}
explode = (0.05, 0.05, 0.05)
A = fig, ax = plt.subplots()
ax.pie(sizes, colors=colors, shadow = True, explode = explode, labels = None, autopct='%1.1f%%',pctdistance=1.25, startangle=90,radius=.8)
plt.legend(labels=labels,loc='lower left', fontsize="8")
plt.title('Are A-Levels Getting Harder?', size=10, weight="bold") 
plt.show(A)
```


> *The proportion of Yes to No answers adds up to 99% when it should add up to 100%. Additionally, visually these proportions should be almost equal but the No proportion of the graph looks like it accounts for almost two-thirds of responses and the Yes responses look like they account for about a third. So visually, the graph is is incorrect. The graph is also missing the one percent one I assume did not agree with yes or no.  Finally, the colors are too similar and need to contrast more for viewers to easily delineate between the categories*
> *I created a graph that included the neither category and showed proper proportionality. Then, I changed the colors to a color-blind-friendly palette.*


## Graph 6
>*The graph is misleading because the y-axis scale does not give units and is not presented in whole numbers, so it is difficult to determine how much the population has grown. Also, regardless of the scale, these seem like very small changes so the scale needs to be adjust so as to make it clear to the reader the size of the these changes. Additionally, the x-axis is also very cluttered, so it is difficult to see correlate the year (x) with the population data (y).*




## Graph 7
```{python}

labels2='Freshman','Sophomore', 'Junior', 'Senior' 
sizes2 = [25, 25, 25, 25]

CB_color_cycle = ['#377eb8', '#ff7f00', '#4daf4a',
                  '#f781bf', '#a65628', '#984ea3',
                  '#999999', '#e41a1c', '#dede00']                          
```

```{python}
explode = (0.05, 0.05, 0.05, 0.05)
C = fig, ax = plt.subplots()
ax.pie(sizes2,colors=colors, shadow = True,explode=explode, labels = None, autopct='%1.1f%%',pctdistance=1.25, startangle=90,radius=1)
plt.legend(labels=labels2,loc='center right', bbox_to_anchor=(1.4, .5), fontsize="8")
plt.title('What Grade Are you In?', size=10, weight="bold") 
plt.show(C)
```

> *The original graph had a number of issues. One, the colors are too similar to distinguish between the groups, Two, the percentages add up to more than 100% which makes zero sense (should always be a portion of 100%). And three, the proportions for the graph are wrong given that they are all close to being equal.*
> *I corrected the proportionality of the segments and the colors. I'm assuming that the 26% from each category is a keystroke error and the original authors meant to put 25%. As you can see, this makes the graph segments equal. Secondly, I auto-assigned color blind friendly colors that contrast well so that all readers can see distinct segments.*

## Graph 8
```{r}
Month <- c("Jan. 2012", "Feb. 2012", "Mar. 2012", "Apr. 2012", "May 2012", "Jun. 2012", "Jul. 2012", "Aug. 2012", "Sep. 2012", "Oct. 2012", "Nov. 2012", "Dec. 2012", "Jan. 2013")
Price <- c(3.17, 3.85, 4.26, 4.15, 4.01, 3.90, 3.60, 3.74, 3.90, 3.40, 3.450, 3.51, 3.57)
PriceLabels <- c("$3.17", "", "", "", "", "", "", "", "", "", "", "$3.51", "$3.57")
GasStuff1 <- data.frame(Month, Price)
GasStuff1$Month <- factor(GasStuff1$Month, levels = GasStuff1$Month)
#GasStuff1
```

```{r}
Gasplot <- ggplot(GasStuff1, aes(x=Month, y=Price,group =1)) + 
  geom_point(color = "#0072B2")+
  geom_line()+
  geom_text(label = PriceLabels, nudge_x=0.1, nudge_y=0.05)+
  theme(axis.text.x=element_text(angle=45,hjust=1))+
  ggtitle("Gas Prices (Jan. 2012 to Jan. 2013)") +
  xlab("Month and Year") + 
  ylab("Gas Prices (USD)")+  
  theme(plot.title = element_text(size=10, hjust=0.5))+  
  theme(axis.title = element_text(size = 10))
  

Gasplot

```


> *The original misleads the viewer in several ways. First, it uses vague time points like "Last year" and "Last week" instead of listing actual dates. Also, it excludes data between time points so it only gives the impression of rising prices while prices may have fallen between the time points. This combined issues skew the data in such a way that it shows continuous increases when there may be data that was omitted that shows a more cyclical nature to the data. They are also mixing between weeks and months when it comes to their time points, so there are some inconsistencies there.*
> *I fixed this by plotting all the data by month over the course of a year. Showing data by month instead of by week also will show 12 data points instead of 52, making the graph more readable while not losing important data. I also highlighted the prices emphasized in the original graph by labeling the data points with the correlated price. I also made the points blue for some level of contrast. As a note, this data is just made up by me, so it may not reflect the actual data. Just showing how this data could be graphed to be more complete and less misleading.*

## Graph 9
```{r}
suppressWarnings(install.packages("plotly", repos = "http://cran.us.r-project.org"))
suppressWarnings(library(plotly))

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
fig <- plot_ly(
    domain = list(x = c(0, 1), y = c(0, 1)),
    value = 634,
    title = list(text = "Your Credit Score"),
    type = "indicator",
    mode = "gauge+number",
      gauge = list(
      axis =list(range = list(300, 850)),
      steps = list (
        list(range = c(300, 579), color = "#D55E00"),
        list(range = c(580, 669), color = "#E69F00"),
        list(range = c(670, 739), color = "#F0E442"),
        list(range = c(740, 799), color = "#40B0A6"),
        list(range = c(800, 850), color = "#009E73")),
      bar = list(color = "#000000")
    
    ))

fig <- fig %>%
  layout(margin = list(l=40,r=40))

fig
```
> *The original graph is misleading for several reasons. One, it has no scaling/proportionality meaning that the unfilled portion of the image should be much smaller since 634 is a significant portion of 700. Similarly, we do not know where this chart begins numerically because there is no scale. Another similar issue is that credit scores go to 850, not 700, so the entire basis of the graph is off. Second, the coloring for the graph is too similar for someone with colorblindness to see the differences. Three, there are no visual delineations about how this score aligns to the score out of 850, making this graph essentially meaningless.*
> *I attempted to correct this by making a gauge plot that ranges from 300 to 850 (the actual range of credit scores) and provided visual delineations. I also added color blind friendly colors. The code I used does not allow for a legend, so I was not able to add what the colors mean, but I think the colors and numbers have a logical flow to indicate that orange is bad and green is good.*



## Graph 10

> *The original misleads the viewer in several ways. First, it doesn't add up to 100%, so it is omitting data. Second, it is just a table so the data is not actually visualized in a meaningful way. Third, It highlights the approval number, drawing the viewer away from the disapproval metric.*
> *I would fix this by creating a colorblind-friendly pie chart that adds a third category called "Neither" to address the missing three percent.*


## Graph 11
```{r}

School <- c('Georgia', 'Alabama', 'Auburn', "Mississippi State", "Florida", "Texas A&M", "Missouri", "South Carolina", "Ole Miss", "LSU", "Arkansas", "Tennessee", "Kentucky", "Vanderbilt")
Projected_Wins <- c(10.9, 10.8, 9.1, 8.4,8.1,7.4,7.3,7.3,6.3,6.2,5.9,5.8,5.4,3.9)
                          
School_FB <- data.frame(School, Projected_Wins)
School_FB <- arrange(School_FB, Projected_Wins)
School_FB$School <- factor(School_FB$School, levels = School_FB$School)

#School_FB

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
suppressWarnings(install.packages("viridis", repo = "http://cran.us.r-project.org"))
suppressWarnings(library(viridis))
color_palette <- viridis(14)

##Creates a color palette with 14 colorblind friendly colors
```


```{r}
ggplot(School_FB, aes(x = School, fill=School, y = Projected_Wins)) +
  geom_col() +
  labs(x = "School Name", y = "Projected Wins", size = 8) +
  ggtitle("Projected Win Totals for College Teams (2018)")+
   geom_text(aes(label = Projected_Wins), hjust = 1.5, vjust = .5, color = "white" )+
theme(legend.position="none")+
  coord_flip()+
  scale_fill_manual(values = color_palette)



```

> *The original graph is not in descending order so it is not easy to immediately determine which team is projected to have the most wins.*
> *I fixed this by creating a horizontal bar chart that lists the schools by the number of projected wins in descending order. I also added a colorblind palette even tho this does not represent actual school colors..*

## Graph 12

> *The original graph is misleading because it shows three different speed gauges that are mathematically similar, but the gauges are not proportional/do not have any scaling/units to provide context per gauge. All three values: Google Chrome (29, 619); Microsoft Edge (31, 786); and Firefox (26, 876) are fairly close to one another, but the middle gauge for Microsoft Edge (31,786) is maxed out/completely full where as Google Chrome (29,619) is only two-thirds full but is only 2,000 less than Microsoft Edge. Clearly, these graphs are not proportional and all graphs should have the same scale such as all of them topping out at 45,000.*
> *Clearly, these graphs are not proportional and all graphs should have the same scale such as all of them topping out at 45,000. Also, coloring needs to be changed so that the gauges are more easily readable.*


## Graph 13
> *This graph is misleading because it doesn't say how much these poverty rates increase (percent change) or adjust for the size of population. Comparing the rate changes in large population countries (UK, Spain, France), to small population countries (Portugal, Sweden, Austria) just doesn't make sense. Unless you account for population by adjusting these metrics for residents per 100k for example, than this is like comparing apples to oranges. This graph also does not convey how big the populations in these age ranges are, another confounding issue. I think the best way to fix this graph to accurately convey the data would be to display line graphs of each country tracking poverty rate changes per age group after adjusting for population. This could all be presented in a number of panels. It may not be as cool as the map, but it would be more accurate.*

##Graph 14


> *Graph 14 suffers from a lack of color diversity, making it difficult to differentiate between the different portions of the graph. I would recommend adding contrasting colors that make it easy for the viewer to immediately delineate between the pie portions.*

## Graph 15
> *This graph suffers from a lack of color and layering issues. Because the graph is in black and white, it is almost impossible to tell the difference between the different lines of the graph. Another issue is that it looks like the graph was layered over the legend, just making the graph look poorly constructed.I would print this in color and double check that the graph is layered correctly to avoid any obstructions of data or lettering.*
