---
title: "Homework 5 - Advanced Data in R"
format: docx
---

```{r}
#| include: false
library(gapminder)
data(gapminder)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

library(dplyr)
library(tidyr)
```

Fill out [this Quarto document](https://github.com/BIFX552-22/CourseInfo/hw/05advanced_data.qmd) while completing the lesson on writing data, `dplyr` and `tidyr`.

### [Writing Data](https://swcarpentry.github.io/r-novice-gapminder/11-writing-data/index.html)

::: callout-tip
## Key Points

-   Save plots from RStudio using the 'Export' button.

-   Use `write.table` to save tabular data.
:::

#### Challenge 1

Modify the code below to print a second page in the pdf, showing a facet plot (hint: use `facet_grid`) of the same data with one panel per continent.

```{r}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)

#page 1

p <- ggplot(gapminder, aes(year, lifeExp, color=country)) +
  geom_line() +
  theme(legend.position = "none")

#page 2

p
p + facet_grid(~continent)

# You then have to make sure to turn off the pdf device!

dev.off()

p
p + facet_grid(~continent)


```

#### Challenge 2

Write a data-cleaning script file that subsets the gapminder data to include only data points collected since 1990.

Use this script to write out the new subset to a file in the `cleaned-data/` directory.

```{r}
ninties_subset <- gapminder[gapminder$year > 1990,]


write.csv(ninties_subset, "nineties.csv")
```

### [Data frames with `dplyr`](https://swcarpentry.github.io/r-novice-gapminder/13-dplyr/index.html)

::: callout-tip
## Key Points

-   Use the `dplyr` package to manipulate data frames.
-   Use `select()` to choose variables from a data frame.
-   Use `filter()` to choose data based on values.
-   Use `group_by()` and `summarize()` to work with subsets of data.
-   Use `mutate()` to create new variables.
:::

#### Challenge 1

Write a single command (which can span multiple lines and includes pipes) that will produce a data frame that has the African values for `lifeExp`, `country` and `year`, but not for other Continents. How many rows does your data frame have and why?

```{r}
# insert code here
africa_data <- filter(gapminder, continent == "Africa") %>%
  select(year, country, lifeExp)

str(africa_data)


#There are 624 rows because there are 52 countries in Africa and each country has 12 rows of data (52 * 12 = 624)
```

#### Challenge 2

Calculate the average life expectancy per country. Which has the longest average life expectancy and which has the shortest average life expectancy?

```{r}
# insert code here
 lifeExp_country <-group_by(gapminder, country) %>%
    summarize(mean_lifeExp = mean(lifeExp ))
 

filter(lifeExp_country, mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))

 

```

#### Challenge 3

Calculate the average life expectancy in 2002 of 2 randomly selected countries for each continent. Then arrange the continent names in reverse order. **Hint:** Use the `dplyr` functions `arrange()` and `sample_n()`, they have similar syntax to other dplyr functions.

```{r}
# insert code here
rando_lifeExp_country <- gapminder %>%
    filter(year == 2002)%>%
    group_by(continent) %>%
    sample_n(2) %>%
    summarize(mean_lifeExp = mean(lifeExp ))%>%
    arrange(desc(mean_lifeExp))

rando_lifeExp_country
```

### [Data frames with `tidyr`](https://swcarpentry.github.io/r-novice-gapminder/14-tidyr/index.html)

::: callout-tip
## Key Points

-   Use the `tidyr` package to change the layout of data frames.

-   Use `pivot_longer()` to go from wide to longer layout.

-   Use `pivot_wider()` to go from long to wider layout.
:::

#### Challenge 1

Is gapminder a purely long, purely wide, or some intermediate format?

Gapminder is some sort of intermediate format. While it is long, it also has multiple columns.

#### gap_wide vs gap_long vs gapminder

The `gap_wide` data set is formatted such that there is only one row per country.

```{r}
library(readr)
gap_wide <- read_csv('https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_wide.csv')

gap_wide
```

The `gap_long` data set is formatted such that there is one row per data point.

```{r}
gap_long <- gap_wide %>%
  pivot_longer(
    cols = c(starts_with('pop'), 
             starts_with('lifeExp'),
             starts_with('gdpPercap')),
    names_to = "obstype_year", 
    values_to = "obs_values") %>%
    
    separate(obstype_year, into = c('obstype', 'year'), sep = '_')

gap_long
```

Both formats cover the extreme ends of the spectrum of how we could format this in a tidy way. The intermediate format provided by `data(gapminder)` is probably more useful in most situations.

```{r}
gapminder
```

#### Challenge 2

Using `gap_long`, calculate the mean life expectancy, population, and gdpPercap for each continent. **Hint:** use the `group_by()` and `summarize()` functions we learned in the `dplyr` lesson

```{r}
long_boy <- gap_long %>% 
  group_by(continent, obstype) %>%
   summarize(means=mean(obs_values))

long_boy

```

#### Challenge 3

Create a `gap_ludicrously_wide` format data by pivoting over countries, year and the 3 metrics? **Hint** this new data frame should only have 5 rows.

```{r}
gap_ludicrously_wide <- gap_long %>%
  group_by(continent)%>%
  unite(id_var, country, obstype, year, sep = "_" ) %>%
  pivot_wider(names_from = id_var, values_from = obs_values)

```

```{r}
gap_normal <- gap_long %>%
    pivot_wider(names_from = obstype, values_from = obs_values) %>%
  select(country, continent, year, lifeExp, pop, gdpPercap) %>%
  arrange(continent, country) %>% 
  
  mutate(year = as.integer(year),
         pop = as.integer(pop))

gap_normal
```

```{r}
new_long <- gap_long %>% 
  unite(id_var, obstype, year, sep = "_") %>%
  pivot_wider(names_from = id_var, values_from = obs_values)

new_long
```
